Individualized Measures of Partisan Fairness and Utility in Redistricting (??)

### 2021-08-29 paper/paper.Rmd
section 2 drafted

---
title: Individual Representation: Harm and Utility(?)
authors:
  - name: Cory McCartan
    thanks: Footnote stuff
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - standing
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    keep_tex: true
---

@@

1.

@@

>>>>

---
title: Individualized Measures of Partisan Fairness and Utility in Redistricting (??)
authors:
  - name: Cory McCartan
    thanks: Footnote stuff
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - standing
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
---

**Title suggestions**

-   Individual Representation: Harm and Utility

-   Redistricting and Individual Representational Harm

-   Fairness and Utility in Redistricting: an Individual Perspective

-   Individualized Measures of Fairness and Utility in Redistricting

@@

The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, individual candidate attributes, and more.
Here, we restrict our attention to what we believe are the two dominant contributors to partisan utility: which party represents a voter's district, and how many overall seats a voter's preferred party wins.

## Notation and Setup

Let $V$ be the set of voters in the region and time period under consideration, who can vote in a set of legislative elections $E$.
Each election involves the selection of one of a set of political parties $P$ to represent each of a set of legislative districts, indexed by $D$.
We take these four sets to be fixed (which does not require that voters do not move, or that parties contest every election).

For each election $e\in E$, every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $\xi_e:V\surjto D$.
The set of all possible districting plans is $\Xi_e$.

Voters either vote for a candidate running in their district, or abstain from voting.
Notationally, we have a set of functions $\vote_e:V\to P\cup\{\textsc{abstain}\}$.
Votes may be tallied at the district level, either in total or by party: \begin{align*}
    \votes_e&: D\to \N, \qquad 
    \votes_e(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v)\in P\} \\
    \votes_e^{(p)}&: D\to \N, \qquad 
    \votes_e^{(p)}(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v) = p\}.
\end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
    \win_e: D\to P, \qquad \win_e(d) = \argmax_{p\in P} \votes_e^{(p)}(d).
$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win_e\circ\xi_e(v)$, and each political party can be said to win a number of seats, formalized as$$
\seats_e:P\to\N, \qquad \seats_e(p) = \sum_{d\in D} \ind\{\win_e(d) = p\}.
$$

## Global Outcome Utility

We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for election $e\in E$ can thus be written$$
u_e^{(G)}:V\to\R, \qquad u_e^{(G)}(v) = h\qty(\frac{\seats_e\circ\vote_e(v)}{|D|}),
$$

for $h:[0,1]\bijto\R$ an increasing function.

What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility?
It depends on the form of $h$, as the following result describes.

\begin{prop}

Let $\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the overall vote.
Define the total utility for this rule as $$
U_e^{(G)}(f) = \sum_{v\in V} h\qty(f\qty(s_{\vote_e(v)})).
$$Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \half\}$, and if $h$ is concave, $U_e^{(G)}$ is maximized by an increasing $f$.
In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.

\end{prop}

Thus, if we are willing to adopt the logarithm as the functional specification of $u_e^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.

## Local Outcome Utility

The specification of the utility of the local election outcome is more straightforward.
We define$$
u_e^{(L)}:V\to\R, \qquad u_e^{(L)}(v) = \ind\{\win_e\circ\xi_e(v) = \vote_e(v)\},
$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.

Attempting to maximize the total local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---must districts be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
The local measure will additionally reward what have been called *bipartisan gerrymanders*, where there are few or no competitive districts.

@@

\appendix

# Proofs of Propositions
